A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-06 paper.tex
Gregor's updates

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.
A necessary first step is a degree of  terminological clarity.
As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as being any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which, because of its origins in an idealised
mathematical model, is more limited in what can be described.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture complexities of family structure which
make, for example, polytomies a necessary feature of local trees.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
were a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner such an  incompehensibly
large space of possibilities.
It is therefore important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here,
is to use polytomies and stacked recombinations as direct topological
descriptions of uncertainty. Other methods to capture, for example,
uncertainty about node ages and breakpoint location, is an important
aspect of future work.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
to downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection statistic~\citep{speidel2019method}
computes selection $p$-values by comparing the rates of coalescence
at various time-points.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
for ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to be to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-free representation that naturally
incorporates uncertainty are well illustrated by the
ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale, with the GISAID database~\citep{shu2017gisaid}
holding over 15 million sequences as of [FIXME].
Recombination, while rare, is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
with recombinants potentially gaining fitness by combining
mutations from deeply diverged lineages [cite].
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 and the other (more sparsely sampled) consisting of
657K sequences sampled up to June 30, 2022.
Sequencing effort at the height of the pandemic produced
tens of thousands of sequences collected in a single day,
with hundreds being identical.
Polytomies are therefore and unescapable reality, and a
fully-resolved binary tree would clearly be false precision.
Because of the longitudinal data collection, there is no
requirement that samples must be leaf nodes, and samples
will frequently descend from other (older) samples.
Unary nodes (a node in a local tree with exactly one child)
are also a common feature, when a sample is inferred to have
exactly one direct descendant.
Despite the abundance of data and the rarity of recombination,
the parental lineages cannot be precisely identified
in many cases. For example, many are inferred to have three
or more parental lineages (in agreement with the
analysis resulting in the Pango designation of the recombinant
lineage); in most cases, these will indicate two or more
independent recombination events.
Faithfully representing the uncertainty
around these events seems a more useful approach than
the arbitrary resolution which would be required,
if encoded as an eARG.

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may help
to clarify inference goals and improve
methods for evaluation.
In the majority of cases,
ARG inference accuracy is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
as originally proposed by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in a indirect manner, and the tree
metrics assume unlabelled nodes, effectively discarding information
about node sharing between trees.
Tree distance metrics usually have $O(n^2)$ time complexity, and so cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance
based approaches, and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which captures many important aspects of dated ARGs
but neglects [SOME OTHER THINGS].
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes, and compared this to the
outputs of several methods.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
% Not sure this is the right place for this sentence
All of these evaluations have been based on
ground truth data from highly idealised simulations
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamendal to the actual representation of the thing]

>>>>

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures. There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees. Flexible ARG encoding should accommodate such cases.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
was a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incompehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here,
is to use polytomies and stacked recombinations as direct topological
descriptions of uncertainty. Other methods to capture, for example,
uncertainty about node ages and breakpoint locations, is an important
aspect of future work.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
to downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection statistic~\citep{speidel2019method}
computes selection $p$-values by comparing the coalescence rates at various time points.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-free representation that naturally
incorporates uncertainty about the ordering of events in and ARG are well illustrated by the
ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale. The GISAID database~\citep{shu2017gisaid}
holds over 15 million sequences as of [FIXME].
Recombination, while rare, is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
with recombinants potentially gaining fitness by combining
mutations from deeply diverged lineages [cite].
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 and the other (more sparsely sampled) consisting of
657K sequences sampled up to June 30, 2022.
Sequencing effort at the height of the pandemic produced
tens of thousands of sequences collected in a single day,
with hundreds being identical.
Polytomies are therefore and unescapable reality, and a
fully-resolved binary tree would clearly be false precision.
Because of longitudinal data collection, there is no
requirement that samples must be leaf nodes, and samples
will frequently descend from other (older) samples.
Unary nodes (a node in a local tree with exactly one child)
are also a common feature, when a sample is inferred to have
exactly one direct descendant.
Despite the abundance of data and the rarity of recombination,
the parental lineages cannot be precisely identified
in many cases. For example, many are inferred to have three
or more parental lineages (in agreement with the
analysis resulting in the Pango designation of the recombinant
lineage); in most cases, these will indicate two or more
independent recombination events.
Faithfully representing the uncertainty
around these events seems a more useful approach than
the arbitrary resolution which would be required,
if encoded as an eARG.

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
the accuracy of ARG inference is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
as originally proposed by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in an indirect manner, and the tree
metrics assume unlabelled nodes, effectively discarding information
about node sharing between trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which captures many important aspects of dated ARGs
but neglects [SOME OTHER THINGS].
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes, and compared this to the
outputs of several methods.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
% Not sure this is the right place for this sentence
All of these evaluations have been based on
ground truth data from highly idealised simulations
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale or multi-generational pedigree datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamendal to the actual representation of the thing]
In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data. We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision. Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.
